\doctitle

### 2023-04-11 msr3.tex
Add appendices; connect MSR3 theory and practice definitions; compiles with NO WARNINGS

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{figures/summary_picture.pdf}
    \caption{Selection of fixed and random effects for LME likelihoods $\mathcal{L}$ using 
    `regularization-agnostic' framework and its SR3 extension using four regularizers. 
    %standard and SR3-relaxed linear mixed-effect likelihoods with popular sparsity-promoting regularizers for the task of simultaneous 
    %The original likelihood $\LL(\beta, \gamma)$ (Eq. \ref{eq:lmm_objective}) is relaxed with SR3 approach to obtain $u(\beta, \gamma)$ (Eq. \ref{eq:value_function_definition}). Both setups are then solved using the same Proximal Gradient Descent (PGD) algorithm. 
    SR3 relaxation accelerates algorithmic converge  (middle panel), and gives better robustness and improved performance on synthetic problems across regularizers (right panel)}.
    \label{fig:summary}
\end{figure}

@@

The negative log likelihood~\eqref{eq:ml_lme_optimization_setup}  is nonlinear and nonconvex, and requires an iterative numerical solver.
However, it is convex with respect to $\beta$, and weakly convex with respect to $\gamma$, with a weak convexity constant $\overline \eta$
computed in~\cite[Section 5.1]{Theory1} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation 
%\eq{
%	\label{eq:beta_formula}
\[
	\beta(\Gamma) = \argmin_{\beta}\LL(\beta, \Gamma) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i.
\] %}
By using the simplification $\Gam=\Diag{\gam}$, we obtain the problem
%{In the meta-analysis setting under study, 
%In our study, we assume that $\Gam$ is a diagonal matrix of the form
%$\Gamma = \Diag{\gamma}, \ \gamma \in \R^s$ with the 
%$\Lambda_i \in \R_{++}^{n_i \times n_i}$ fixed and given 
%%the positive definite constraint from (\ref{eq:ml_lme_optimization_setup}) transforms into a box constraint:
%so that our problem takes the form}
\eq{
	\label{eq:lme_diagonal_setup}
	\min_{\beta\in\R^p, \gam\in\R^q_+} & \ \LL(\beta,\gam)
	:=\LL_{ML}(\beta, \Diag{\gamma}) 
%	\\
%	s.t. & \ \gamma \geq 0\ .
	}
%Under the assumption $\Lambda_i \in \R_{++}^{n_i \times n_i}$ in~\eqref{eq:lme_setup}, 
In this setting, when an
entry $\gamma_j$ takes the  value $0$ the corresponding coordinates of all random effects $u_{ij}$ are identically $0$ for all $i$.

@@

 \begin{figure}[H]
    \centering
    \begin{tabular}{ccc}
        \includegraphics[width=0.3\textwidth]{figures/l1_regularizer.pdf}
&
           \includegraphics[width=0.3\textwidth]{figures/lh_regularizer.pdf}
    &
    \includegraphics[width=0.3\textwidth]{figures/l0_regularizer.pdf}\\
  %      \includegraphics[width=0.3\textwidth]{figs/scad_regularizer.pdf}\\
              (a) $\ell_1$  & (b) $\ell_{1/2}$ & (c) $\ell_0$
    \end{tabular}
    \caption{Common convex and non-convex regularizers used for feature selection.}
    \label{fig:regularizers}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{figures/intuition_current.pdf}
    \caption{Proximal Gradient Descent (PGD) for~\eqref{eq:value_function_optimization} converges far faster than for~\eqref{eq:lme_loss_original_in_x}, because the SR3-value function $u_\eta$ yields more spherical level-sets than original likelihood for both convex components $\beta$ (first row) and non-convex components $\gamma$ (second row). 
    %The effect is particularly strong for $\gamma$-minimizers near the boundary of the constraint region (bottom-left panel).
    }
    \label{fig:geometric_intuition_sr3}
\end{figure}

@@

To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we define a value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
\eq{
		u_\eta(w) & = \min_x \ \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w)\\
		S_\eta(w) & = \argmin_x \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w).
}
Substituting the above into (\ref{eq:lme_loss_sr3}) the problem transforms into the problem of optimizing a regularized value function:
\eq{
	\label{eq:value_function_optimization}
	\min_{w} u_\eta(w) + R(w)
}
%Thus at a conceptual level, 
Here we have transformed the original regularized likelihood~\eqref{eq:lme_loss_original_in_x}  %has been transformed 
through relaxation and partial 
minimization to obtain an equivalent problem~\eqref{eq:value_function_optimization} 
for $w$ with the same regularizer. The value function $u_\eta$ encapsulates 
global variational information on the function $\LL(x) + \delta_{\CC}(x)$
relative to $w$.
%Problem~\eqref{eq:value_function_optimization} 
%attempts to find sparse $w$ that are small with respect to $u_\eta$.

@@

In order to solve for the $x^+$ update in line 2 of Algorithm~\ref{alg:pgd_for_value}, we must optimize a convex loss with linear inequality constraints, that is, 
for a fixed $w = (\hat \beta, \hat \gamma)$, we need to solve 
\begin{equation}
\label{eq:ineqLoss}
\min_{\beta, \gamma} \mathcal{L}(\beta, \gamma) + \kappa_{\eta}(\beta - \hat \beta, \gamma - \hat \gamma) \quad \mbox{s.t.} \quad 0 \leq \gamma. 
\end{equation}
This problem is well suited for an interior point approach~\citep{Kojima1991,Nesterov1994,Wright1997}. 
%We now describe how to apply IP methods \teal{to solve the inner problem~\eqref{eq:lme_loss_relaxed} with respect to $x = (\beta, \gamma)$, 
%evaluating the value function $u(w)$ in~\eqref{eq:value_function_definition}}. 
%\eqref{eq:lme_loss_original}. 
% when $\LL=\LL_{ML}$. The case where $\LL=\LL_{REML}$ follows a 
% similar pattern.
%To apply IP methods to our problem class, we first 
First, the inequality constraint $0\le \gamma$ is relaxed using a log-barrier penalty, obtaining a minimization problem for a relaxed objective $\mathcal{L}_{\mu,\eta}$: 
\eq{
	\label{eq:log_barrier_relaxation_for_ip}
	\min_{\beta, \gamma} &\left\{\mathcal{L}_{\mu,\eta}(\beta, \gamma) :=  \mathcal{L}(\beta, \gamma) + \kappa_{\eta}(\beta - \hat \beta, \gamma - \hat \gamma) - \mu\sum_{i=1}^{q}\ln(\gamma_i) \right\}.
}
Here the log-barrier penalty approximates the indicator function to the positive orthant as $\mu$ decreases; indeed, the function $\gamma\mapsto\mu\ln(\gamma)$ epi-converges to the indicator function $\delta_{\mathbb{R}^n_+}(\gamma)$ as $\mu \downarrow 0$ (\cite{rockafellar2009variational}). The penalty (homotopy) parameter $\mu$ is progressively decreased to $0$ as the algorithm proceeds as described below. 
%in the limit minimizing $\mathcal{L}_{\eta,\mu}$ gives the right $x^+$ update in Algorithm~\ref{alg:pgd_for_value}. 
The existence of solutions for the problem~(\ref{eq:log_barrier_relaxation_for_ip}) for any positive $\mu$ is shown in \cite[Theorem 5]{Theory1}, and the convergence of solutions to the MSR3 solution as $\mu\downarrow 0$ is shown in~\cite[Theorem 7]{Theory1}.  
Finally, \cite[Theorem 6]{Theory1} shows that the MSR3 relaxation is consistent with respect to the barrier, so that as the MSR3 parameter $\eta \uparrow \infty$,
 limit points of global solutions to the former are global solutions to the latter. 
 However, in the applications considered here,
 the empirical studies in Sections \ref{sec:synthetic} and \ref{sec:real}
 indicate that one does not need to make $\eta$ particularly large in order to accurately identify the correct sparsity pattern.

%The Lagrangian for~\eqref{eq:log_barrier_relaxation_for_ip} is obtained by dualizing the inequality $\gamma \geq 0$ constraint: 
% Define the objective for \eqref{eq:log_barrier_relaxation_for_ip} by
% \eq{
% 	F_\mu(\beta, \gamma) & = \LL(\beta, \gamma) +  \frac{\lambda_b}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_g}{2}\|\gamma - \tgamma^{(k)}\|_2^2 - \mu\sum_{i=1}^{q}\log(\gamma_i) %+ v^T\gamma \\
% }

For $\gamma>0$, the necessary optimality conditions 
for $\mathcal{L}_{\mu,\eta}$ in $\gamma$ give us the relation 
%between dual variables $v$ and the primal variables $\gamma$: 
\eq{
	\nabla_\gamma \mathcal{L}_{\mu,\eta}(\beta,\gamma) = 
	%\begin{bmatrix}
\nabla_\gam \LL(\beta,\gam)+\eta (\gamma - \hat\gamma)
-\mu \Diag{\gam}^{-1}\one
=0,
%		v_1 - \mu/\gamma_1 \\
%		\dots \\
%		v_q - \mu/\gamma_q 
%	\end{bmatrix} = \begin{bmatrix}
%		0 \\
%		\dots \\
%		0 
%	\end{bmatrix} 
%	\implies v_i\gamma_i = \mu \text{ for } i=1,\dots,q.
}
where $\one$ is the vector of all ones of the 
appropriate dimension.
By setting 
\[
v=\nabla_\gam \mathcal{L}_{\mu,\eta}(\beta,\gam)+\eta(\gamma - \hat \gamma),
\] 
we can rewrite this equation as
%This equation can be written in a matrix form as 
\eq{
	\label{eq:complementary_slackness_kkt}
	v\odot\gamma - \mu\one = 0,
}
where $\one$ is the vector of all ones of the appropriate dimension and
``$\odot$'' denotes the Hadamard (or simply element-wise) product.
%Together with the remaining optimality conditions, we obtain a set of nonlinear equations  that form the the KKT system for~\eqref{eq:log_barrier_relaxation_for_ip}:
The complete set of optimality conditions for 
\eqref{eq:log_barrier_relaxation_for_ip} can now be written as
\eq{\label{eq:IP equations}
	G_{\mu,\eta}(v, \beta, \gamma) & 
%	= \begin{bmatrix}
%		\nabla_v F_\mu \\
%		\nabla_\beta F_\mu \\
%		\nabla_\gamma F_\mu 
%	\end{bmatrix} 
:= \begin{bmatrix}
v\odot \gamma - \mu\one \\
\nabla_\beta \LL(\beta, \gamma) + \eta(\beta - \hat \beta) \\
\nabla_\gamma \LL(\beta, \gamma) + \eta(\gamma - \hat \gamma) - v
\end{bmatrix} = 0.
}
We then apply Newton's method to~\eqref{eq:IP equations}, 
that is, in each iteration the search direction $[\Delta v, \Delta \beta, \Delta \gamma]$ solves the linear system
\eq{
	\label{eq:ip_iteration_rs_form}
	\nabla G_{\mu,\eta}(v, \beta, \gamma)\begin{bmatrix}
		\Delta v \\
		\Delta \beta \\
		\Delta \gamma
	\end{bmatrix} = -G_{\mu,\eta}(v, \beta, \gamma).
}
where 
\eq{
	\nabla G_{\mu, \eta}(v, \beta, \gamma) = \begin{bmatrix}
 		\Diag{\gamma} & 0 & \Diag{v} \\
 		0 & \nabla^2_{\beta\beta}\LL + \eta I & \nabla^2_{\beta\gamma} \LL\\
 		-I & \nabla^2_{\gamma\beta}\LL & \nabla^2_{\gamma\gamma}\LL + (\eta + \overline{\lambda})	 I
 	\end{bmatrix}
}
and we have used the fact that  $v\odot \gamma=\Diag{v}\gam=\Diag{\gam}v$. 
The exact formulae for the derivatives of $\LL$ are provided in the Appendix~\ref{appendix:derivatives_of_lmm}.

The general structure of the algorithm is as follows.
Given a search direction
$[\Delta v^{(k)}, \Delta \beta^{(k)}, \Delta \gamma^{(k)}]$, 
choose a step of size $\alpha_k>0$
%\eq{
%    \label{eq:ip_step}
%	v^{(k+1)} & = v^{(k)} + \alpha_k \Delta v \\
%	\gamma^{(k+1)} & = \gamma^{(k)} + \alpha_k \Delta\gamma \\
%	\beta^{(k+1)} & = \beta^{(k)} + \alpha_k \Delta \beta
%}
%with $\alpha_k$ 
so that the update
\[
\begin{pmatrix}
v^{(k+1)}\\ \beta^{(k+1)}\\ \gamma^{(k+1)}
\end{pmatrix}
=
\begin{pmatrix}
v^{(k)}\\ \beta^{(k)}\\ \gamma^{(k)}
\end{pmatrix}
+\alf_k
\begin{pmatrix}
\Delta v^{(k)}\\ \Delta\beta^{(k)}\\ \Delta\gamma^{(k)}
\end{pmatrix}
\]
satisfies the conditions
% chosen to satisfy positivity and sufficient descent conditions:
\eq{
    \label{eq:ip_steplen_conditions}
	\text{\textit{Positivity:}} \qquad & \gamma^{(k+1)} > 0,\ v^{(k+1)} > 0 \\
	\text{\textit{Sufficient Descent:}}\qquad & \| G_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|G_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| .
}
At each iteration the relaxation parameter $\mu$ is updated by the formula 
\eq{
    \label{eq:ip_mu_update}
    \mu^{(k+1)} = {v^{(k)}}^T\gamma^{(k)}/q,
}
where ${v^{(k)}}^T\gamma^{(k)}$ is the duality gap at
iteration $k$. The algorithm terminates when the criteria 
\eq{
\label{eq:ip_convergence_criterion}
\begin{aligned}
\|G_{\mu,\eta}(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| &\leq \text{\texttt{tol}}\\
\mu &\leq  \text{\texttt{tol}} 
\end{aligned}
}
are both satisfied, so the interior point problem is nearly stationary, and closely approximates the original problem~\eqref{eq:ineqLoss}.
MSR3 is summarized in Algorithm~\ref{alg:MSR3}, which approximates Algorithm~\ref{alg:pgd_for_value} as the tolerance goes to $0$. 
In the numerical experiments, we use $\text{\texttt{tol}} = 10^{-5}$, and accuracy does not change as the tolerance parameter decreases.

\begin{algorithm}[H]
\SetAlgoLined
$w = w_0$ \\
 \While{not converged}{
    $x^+ \; \mbox{satisfies }\;  \|G_{\mu,\eta}(v^+, x^+)\| \leq \text{\texttt{tol}}, \; \mu \leq \text{\texttt{tol}} $\\
    $w^+ = \prox_{\alpha^{-1} R}(x^+)$
 }
 \caption{\label{alg:MSR3} MSR3}
\end{algorithm}

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/summary_picture.pdf}
    \caption{Selection of fixed and random effects for LME likelihoods $\mathcal{L}$ using 
    `regularization-agnostic' framework and its SR3 extension using four regularizers. 
    %standard and SR3-relaxed linear mixed-effect likelihoods with popular sparsity-promoting regularizers for the task of simultaneous 
    %The original likelihood $\LL(\beta, \gamma)$ (Eq. \ref{eq:lmm_objective}) is relaxed with SR3 approach to obtain $u(\beta, \gamma)$ (Eq. \ref{eq:value_function_definition}). Both setups are then solved using the same Proximal Gradient Descent (PGD) algorithm. 
    SR3 relaxation accelerates algorithmic converge  (middle panel), and gives better robustness and improved performance on synthetic problems across regularizers (right panel)}.
    \label{fig:summary}
\end{figure}

@@

The negative log likelihood~\eqref{eq:ml_lme_optimization_setup}  is nonlinear and nonconvex, and requires an iterative numerical solver.
However, it is convex with respect to $\beta$, and weakly convex with respect to $\gamma$, with a weak convexity constant $\overline \eta$
computed in~\cite[Section 5.1]{Theory1} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation 
%\eq{
%	
\[
\label{eq:beta_formula}
	\beta(\Gamma) = \argmin_{\beta}\LL(\beta, \Gamma) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i.
\] %}
By using the simplification $\Gam=\Diag{\gam}$, we obtain the problem
%{In the meta-analysis setting under study, 
%In our study, we assume that $\Gam$ is a diagonal matrix of the form
%$\Gamma = \Diag{\gamma}, \ \gamma \in \R^s$ with the 
%$\Lambda_i \in \R_{++}^{n_i \times n_i}$ fixed and given 
%%the positive definite constraint from (\ref{eq:ml_lme_optimization_setup}) transforms into a box constraint:
%so that our problem takes the form}
\eq{
	\label{eq:lme_diagonal_setup}
	\min_{\beta\in\R^p, \gam\in\R^q_+} & \ \LL(\beta,\gam)
	:=\LL_{ML}(\beta, \Diag{\gamma}) 
%	\\
%	s.t. & \ \gamma \geq 0\ .
	}
%Under the assumption $\Lambda_i \in \R_{++}^{n_i \times n_i}$ in~\eqref{eq:lme_setup}, 
In this setting, when an
entry $\gamma_j$ takes the  value $0$ the corresponding coordinates of all random effects $u_{ij}$ are identically $0$ for all $i$.

@@

 \begin{figure}[t]
    \centering
    \begin{tabular}{ccc}
        \includegraphics[width=0.3\textwidth]{figures/l1_regularizer.pdf}
&
           \includegraphics[width=0.3\textwidth]{figures/lh_regularizer.pdf}
    &
    \includegraphics[width=0.3\textwidth]{figures/l0_regularizer.pdf}\\
  %      \includegraphics[width=0.3\textwidth]{figs/scad_regularizer.pdf}\\
              (a) $\ell_1$  & (b) $\ell_{1/2}$ & (c) $\ell_0$
    \end{tabular}
    \caption{Common convex and non-convex regularizers used for feature selection.}
    \label{fig:regularizers}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/intuition_current.pdf}
    \caption{Proximal Gradient Descent (PGD) for~\eqref{eq:value_function_optimization} converges far faster than for~\eqref{eq:lme_loss_original_in_x}, because the SR3-value function $u_\eta$ yields more spherical level-sets than original likelihood for both convex components $\beta$ (first row) and non-convex components $\gamma$ (second row). 
    %The effect is particularly strong for $\gamma$-minimizers near the boundary of the constraint region (bottom-left panel).
    }
    \label{fig:geometric_intuition_sr3}
\end{figure}

@@

To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we, for this section alone, define a simplified version of the value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
\eq{
	\label{eq:value_function_definition_only_eta}
		u_\eta(w) & = \min_x \ \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w)\\
		S_\eta(w) & = \argmin_x \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w).
}
Substituting the above into (\ref{eq:lme_loss_sr3}) the problem transforms into the problem of optimizing a regularized value function:
\eq{
	\label{eq:value_function_optimization}
	\min_{w} u_\eta(w) + R(w)
}
%Thus at a conceptual level, 
Here we have transformed the original regularized likelihood~\eqref{eq:lme_loss_original_in_x}  %has been transformed 
through relaxation and partial 
minimization to obtain an equivalent problem~\eqref{eq:value_function_optimization} 
for $w$ with the same regularizer. The value function $u_\eta$ encapsulates 
global variational information on the function $\LL(x) + \delta_{\CC}(x)$
relative to $w$.
%Problem~\eqref{eq:value_function_optimization} 
%attempts to find sparse $w$ that are small with respect to $u_\eta$.

@@

In order to solve for the $x^+$ update in line 2 of Algorithm~\ref{alg:pgd_for_value}, we must optimize a convex loss with linear inequality constraints, that is, 
for a fixed $w = (\hat \beta, \hat \gamma)$, we need to solve 
\begin{equation}
\label{eq:ineqLoss}
\min_{\beta, \gamma} \mathcal{L}(\beta, \gamma) + \kappa_{\eta}(\beta - \hat \beta, \gamma - \hat \gamma) \quad \mbox{s.t.} \quad 0 \leq \gamma. 
\end{equation}
This problem is well suited for an interior point approach~\citep{Kojima1991,Nesterov1994,Wright1997}. 
%We now describe how to apply IP methods \teal{to solve the inner problem~\eqref{eq:lme_loss_relaxed} with respect to $x = (\beta, \gamma)$, 
%evaluating the value function $u(w)$ in~\eqref{eq:value_function_definition}}. 
%\eqref{eq:lme_loss_original}. 
% when $\LL=\LL_{ML}$. The case where $\LL=\LL_{REML}$ follows a 
% similar pattern.
%To apply IP methods to our problem class, we first 
First, the inequality constraint $0\le\gam$ is relaxed
using the perspective of the negative log, i.e %log-barrier function
$\map{\vphi}{\R^q\times\R}{\R\cup\{\infty\}}$ given by
\[
\vphi(\gam,\mu):=
\begin{cases}
-\mu\sum_{i=1}^q\ln (\gam_i/\mu)&,\ \mu>0,\\
\del_{\R^q_+}(\gam)&,\ \mu=0,\\
+\infty&,\ \mu<0.
\end{cases}
\]
The mapping $\vphi$ is known to be a closed proper convex function and, 
for $\mu>0$, it is
essentially equivalent to the well-known log-barrier function.
%since in this case
%\[
%\phi_\mu(\gam)=q\mu\ln(\mu)-\mu\sum_{i=1}^q\ln (\gam_i).
%\]
For more information on the perspective mapping, its calculus,
and perspective duality, we refer the reader to \cite{ABDFM18,ABF13}.
%\red{Should we mention the perspective of $\psi(\gam)=-\ln (\gam)$, or perhaps the closure of the epi-limit? The connection to the perspective map might be useful since
%we get the joint convexity in $\mu$ and $\gam$.}
We call $\eta$ the coupling parameter and $\mu$ the log-barrier parameter
and write $\phimu(\cdot):=\vphi(\cdot,\mu)$.
%The parameter $0\le\blam$ is fixed and is related to the weak convexity of $\LL$
%discussed in Lemma \ref{lem:LL weak cvx}.
%}
The relaxed problem employs 
auxiliary variables $\tbg$ 
and relaxation parameters $0\le\eta$ and $0\le \mu$
to obtain the problem 
\begin{equation}\label{eq:relax} 
\begin{aligned}%\eq{\label{eq:lme_loss_relaxed}
%\nu_\emu:=	
\min_{\bg, \tbg} & \LL(\beta, \gamma) 
	+\phi_\mu(\gam)+\kappa_\eta(\beta-\tbeta,\gam-\tgam)
%\frac{\eta}{2}\norm{\beta - \tbeta}^2 + \frac{\blam+\eta}{2}\norm{\gamma - \tgamma}^2  
+ R(\tbeta, \tgamma) 
\\
\text{s.t. } & \tgamma \geq 0\ .
\end{aligned}
\end{equation}

We rewrite \eqref{eq:relax} %\eqref{eq:lme_loss_relaxed} 
so as to separate the smooth and nonsmooth components to obtain
\eq{ \label{eq:relax2}
    \min_{\bg, \tbg} & \LL_{\emu}(\bg,\tbg) + R\tbg +\del_{\R^q_+}(\tgam)\ ,
%    \\
%    \text{s.t. } & 0\le\tgam,
}
where 
%$x=(\beta, \gamma)$, 
%$w=(\tbeta, \tgamma)$, $\CC=\R^p\times\R^q_+$, and
%$\LL_\eta$ is defined in \eqref{eq:defn L sub eta}.
\eq{\label{eq:LLlam}
    \LL_{\emu}(\bg, \tbg) := 
    \LL\bg + \phimu(\gam)+\keta(\beta - \tbeta, \gamma - \tgamma).
}
Observe that, for all $\mu,\eta\in\R_+$, $\LL_{\emu},\ \nabla \LL_{\emu}$ and $\nabla^2 \LL_{\emu}$ are continuous
on $(\Rp\times\dom(\phimu))\times(\Rp\times\Rq)$ 
(see Appendix \ref{appendix:derivatives_of_lmm}) 
so that $\LL_{\emu}$ is smooth on its domain.
As in \cite{Zheng2019SR3}, we use the 
decoupling to write \eqref{eq:relax2}
as an iterated optimization problem over the smooth components of the objective.
This yields a representation of the form
\begin{equation}
    \label{eq:relax3}
    \min_{\tbg} \uemu\tbg + R\tbg+\del_{\Rqp}(\tgam),
\end{equation}
where %$w=(\tbeta,\tgam)$ and
\begin{equation}
    \label{eq:value_function_definition}
    \uemu\tbg := \min_{\bg } 
    \LL_{\emu}(\bg, \tbg)\ .
%    \hLL(\bg,\tgam)+ %\del_{\Rqp}(\gam)+
%    \frac{\eta}{2}\norm{\bg-\tbg}^2
    %\LL(x) + \lambda\|x - w\|_2^2 
\end{equation}
This is the formulation of the mixed-effects variable selection problem we study. 
Notice that this value function differs from a simplified 
definition in \eqref{eq:value_function_definition_only_eta}: we replaced 
the linearity constraint with a perspective function and got another parameter $\mu$ due to it. The log-barrier penalty approximates the indicator function to the positive orthant as $\mu$ decreases; indeed, the function $\gamma\mapsto\mu\ln(\gamma)$ epi-converges to the indicator function $\delta_{\mathbb{R}^n_+}(\gamma)$ as $\mu \downarrow 0$ (\cite{rockafellar2009variational}). The penalty (homotopy) parameter $\mu$ is progressively decreased to $0$ as the algorithm proceeds as described below. 
%in the limit minimizing $\mathcal{L}_{\eta,\mu}$ gives the right $x^+$ update in Algorithm~\ref{alg:pgd_for_value}. 
 We refer to \eqref{eq:value_function_definition} when we say ``value function'' from now on.

Our focus is on the optimal value function $\uemu$ which captures global 
variational information about the function $\LL$ over its domain.
We show that $\uemu$ has a locally Lipschitz continuous gradient
and that the evaluation of $\uemu$ and $\nabla \uemu$ is accomplished
by optimizing a well conditioned strongly convex function. This
allows us to apply the PGD algorithm to the function $\uemu$ rather than
the function $\LL$. Our numerical studies show that the global information 
captured by $\uemu$ significantly improves both the accuracy of the solution
obtained and the overall numerical efficiency of the algorithm.

To motivate the reader, the theoretical results that support the design of Algorithm \ref{alg:MSR3} go as follows. The existence of solutions for the problem~(\ref{eq:value_function_definition}) for any positive $\mu$ is shown in \cite[Theorem 5]{Theory1}, and the convergence of solutions to the MSR3 solution as $\mu\downarrow 0$ is shown in~\cite[Theorem 7]{Theory1}.  
Finally, \cite[Theorem 6]{Theory1} shows that the MSR3 relaxation is consistent with respect to the barrier, so that as the MSR3 parameter $\eta \uparrow \infty$,
 limit points of global solutions to the former are global solutions to the latter. 
 However, the empirical studies in Sections \ref{sec:synthetic} and \ref{sec:real}
 indicate that one does not need to make $\eta$ particularly large in order to accurately identify the correct sparsity pattern. These results were first published in the joint works \cite{sholokhov2022relaxation, aravkin2022jimtheory}, constitute the original work of the author of the thesis, and were mostly taken verbatim per the co-authors'  permission.

%The Lagrangian for~\eqref{eq:value_function_definition} is obtained by dualizing the inequality $\gamma \geq 0$ constraint: 
% Define the objective for \eqref{eq:value_function_definition} by
% \eq{
% 	F_\mu(\beta, \gamma) & = \LL(\beta, \gamma) +  \frac{\lambda_b}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_g}{2}\|\gamma - \tgamma^{(k)}\|_2^2 - \mu\sum_{i=1}^{q}\log(\gamma_i) %+ v^T\gamma \\
% }

For $\gamma>0$, the necessary optimality conditions 
for $\mathcal{L}_{\mu,\eta}$ in $\gamma$ give us the relation 
%between dual variables $v$ and the primal variables $\gamma$: 
\eq{
	\nabla_\gamma \mathcal{L}_{\mu,\eta}(\beta,\gamma) = 
	%\begin{bmatrix}
\nabla_\gam \LL(\beta,\gam)+\eta (\gamma - \hat\gamma)
-\mu \Diag{\gam}^{-1}\one
=0,
%		v_1 - \mu/\gamma_1 \\
%		\dots \\
%		v_q - \mu/\gamma_q 
%	\end{bmatrix} = \begin{bmatrix}
%		0 \\
%		\dots \\
%		0 
%	\end{bmatrix} 
%	\implies v_i\gamma_i = \mu \text{ for } i=1,\dots,q.
}
where $\one$ is the vector of all ones of the 
appropriate dimension.
By setting 
\[
v=\nabla_\gam \mathcal{L}_{\mu,\eta}(\beta,\gam)+\eta(\gamma - \hat \gamma),
\] 
we can rewrite this equation as
%This equation can be written in a matrix form as 
\eq{
	\label{eq:complementary_slackness_kkt}
	v\odot\gamma - \mu\one = 0,
}
where $\one$ is the vector of all ones of the appropriate dimension and
``$\odot$'' denotes the Hadamard (or simply element-wise) product.
%Together with the remaining optimality conditions, we obtain a set of nonlinear equations  that form the the KKT system for~\eqref{eq:value_function_definition}:
The complete set of optimality conditions for 
\eqref{eq:value_function_definition} can now be written as
\eq{\label{eq:IP equations}
	G_{\mu,\eta}(v, \beta, \gamma) & 
%	= \begin{bmatrix}
%		\nabla_v F_\mu \\
%		\nabla_\beta F_\mu \\
%		\nabla_\gamma F_\mu 
%	\end{bmatrix} 
:= \begin{bmatrix}
v\odot \gamma - \mu\one \\
\nabla_\beta \LL(\beta, \gamma) + \eta(\beta - \hat \beta) \\
\nabla_\gamma \LL(\beta, \gamma) + \eta(\gamma - \hat \gamma) - v
\end{bmatrix} = 0.
}
We then apply Newton's method to~\eqref{eq:IP equations}, 
that is, in each iteration the search direction $[\Delta v, \Delta \beta, \Delta \gamma]$ solves the linear system
\eq{
	\label{eq:ip_iteration_rs_form}
	\nabla G_{\mu,\eta}(v, \beta, \gamma)\begin{bmatrix}
		\Delta v \\
		\Delta \beta \\
		\Delta \gamma
	\end{bmatrix} = -G_{\mu,\eta}(v, \beta, \gamma).
}
where 
\eq{
	\nabla G_{\mu, \eta}(v, \beta, \gamma) = \begin{bmatrix}
 		\Diag{\gamma} & 0 & \Diag{v} \\
 		0 & \nabla^2_{\beta\beta}\LL + \eta I & \nabla^2_{\beta\gamma} \LL\\
 		-I & \nabla^2_{\gamma\beta}\LL & \nabla^2_{\gamma\gamma}\LL + (\eta + \overline{\lambda})	 I
 	\end{bmatrix}
}
and we have used the fact that  $v\odot \gamma=\Diag{v}\gam=\Diag{\gam}v$. 
The exact formulae for the derivatives of $\LL$ are provided in the Appendix~\ref{appendix:derivatives_of_lmm}.

The general structure of the algorithm is as follows.
Given a search direction
$[\Delta v^{(k)}, \Delta \beta^{(k)}, \Delta \gamma^{(k)}]$, 
choose a step of size $\alpha_k>0$
%\eq{
%    \label{eq:ip_step}
%	v^{(k+1)} & = v^{(k)} + \alpha_k \Delta v \\
%	\gamma^{(k+1)} & = \gamma^{(k)} + \alpha_k \Delta\gamma \\
%	\beta^{(k+1)} & = \beta^{(k)} + \alpha_k \Delta \beta
%}
%with $\alpha_k$ 
so that the update
\eq{
\label{eq:ip_one_step}
\begin{pmatrix}
v^{(k+1)}\\ \beta^{(k+1)}\\ \gamma^{(k+1)}
\end{pmatrix}
=
\begin{pmatrix}
v^{(k)}\\ \beta^{(k)}\\ \gamma^{(k)}
\end{pmatrix}
+\alf_k
\begin{pmatrix}
\Delta v^{(k)}\\ \Delta\beta^{(k)}\\ \Delta\gamma^{(k)}
\end{pmatrix}
}
satisfies the conditions
% chosen to satisfy positivity and sufficient descent conditions:
\eq{
    \label{eq:ip_steplen_conditions}
	\text{\textit{Positivity:}} \qquad & \gamma^{(k+1)} > 0,\ v^{(k+1)} > 0 \\
	\text{\textit{Sufficient Descent:}}\qquad & \| G_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|G_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| .
}
At each iteration the relaxation parameter $\mu$ is updated by the formula 
\eq{
    \label{eq:ip_mu_update}
    \mu^{(k+1)} = {v^{(k)}}^T\gamma^{(k)}/q,
}
where ${v^{(k)}}^T\gamma^{(k)}$ is the duality gap at
iteration $k$. The algorithm terminates when the criteria 
\eq{
\label{eq:ip_convergence_criterion}
\begin{aligned}
\|G_{\mu,\eta}(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| &\leq \text{\texttt{tol}}\\
\mu &\leq  \text{\texttt{tol}} 
\end{aligned}
}
are both satisfied, so the interior point problem is nearly stationary, and closely approximates the original problem~\eqref{eq:ineqLoss}.
MSR3 is summarized in Algorithm~\ref{alg:MSR3}, which approximates Algorithm~\ref{alg:pgd_for_value} as the tolerance goes to $0$. 
In the numerical experiments, we use $\text{\texttt{tol}} = 10^{-5}$, and accuracy does not change as the tolerance parameter decreases.

\begin{algorithm}[H]
\SetAlgoLined
$w^0 = (\tbeta^0, \tgamma^0)$, $x^0 = (\beta^0, \gamma^0)$, $\gammax>\tgam^0$, $\alpha < \frac{2}{L}$ 
for $L>0$.\\
 \While{not converged and $\tgam\le\gammax$}{
    \text{Find } $x^+$ \text{ using \eqref{eq:ip_one_step}}$\mbox{ which satisfies }\;  \|G_{\mu,\eta}(v^+, x^+)\| \leq \text{\texttt{tol}}, \; \mu \leq \text{\texttt{tol}} $\\
    $w^+ = \prox_{\alpha^{-1} R}(x^+)$
 }
\caption{\label{alg:MSR3} MSR3}
\end{algorithm}
\medskip
